---
title: "AlberdiLab | van-Riemsdijk et al. in prep"
subtitle: "Study title to be added"
data analysis:
  - Ostaizka Aizpurua^[University of Copenhagen, ostaizka.aizpurua@sund.ku.dk]
  - Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://alberdilab.github.io/bombina_hologenomics
description: |
  Data analysis code for the study on the recovery of metagenome‑assembled genomes and derived microbial communities from lizard faecal samples.
link-citations: yes
github-repo: alberdilab/bombina_hologenomics
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for data analysis in study of the population-level metagenomic data of Podarcis muralis lizards across elevational gradients in various mountain ranges of the Pyrenees.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/bombina_hologenomics.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=FALSE, comments="", message=FALSE}
# Base
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(tinytable)
library(rairtable)

# For tree handling
library(ape)
library(phyloseq)
library(phytools)

# For plotting
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(ggnewscale)
library(gridExtra)
library(ggtreeExtra)
library(ggtree)
library(ggh4x)

# For statistics
library(spaa)
library(vegan)
library(Rtsne)
library(geiger)
library(hilldiv2)
library(distillR)
library(broom.mixed)
#library(lmerTest)
library(Hmsc)
library(corrplot)
library(ANCOMBC)
```

<!--chapter:end:index.Rmd-->

# Data preparation

## Load data

Load the original data files outputted by the bioinformatic pipeline.

### Sample metadata

```{r load_sample_metadata, warning=FALSE, comments="", message=FALSE}
sample_metadata <- read_tsv("data/DMB0163_metadata.tsv.gz") %>%
    rename(sample=1)
```

### Genome metadata

```{r load_genome_metadata, warning=FALSE, comments="", message=FALSE}
genome_metadata <- read_tsv("data/DMB0163_mag_info.tsv.gz") %>%
    rename(length=mag_size)
```

### Read counts

```{r load_read_counts, warning=FALSE, comments="", message=FALSE}
read_counts <- read_tsv("data/DMB0163_counts.tsv.gz") %>%
    rename(genome=1) %>%
    select(all_of(c("genome",sample_metadata$sample))) %>% 
    arrange(match(genome,genome_metadata$genome))
```

### Genome base hits

```{r load_genome_hits, warning=FALSE, comments="", message=FALSE}
genome_coverage <- read_tsv("data/DMB0163_coverage.tsv.gz") %>%
    rename(genome=1) %>%
    select(all_of(c("genome",sample_metadata$sample))) %>% # sort samples
    arrange(match(genome,genome_metadata$genome)) # sort genomes
```

### Genome tree

```{r load_genome_tree, warning=FALSE, comments="", message=FALSE}
genome_tree <- read_tree("data/DMB0163.tree")
genome_tree$tip.label <- str_replace_all(genome_tree$tip.label,"'", "") #remove single quotes in MAG names
genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome) # keep only MAG tips
```
### Genome annotations

Downloading individual annotation files from ERDA using information in Airtable and writing them to a single compressed table takes a while. The following chunk only needs to be run once, to generate the ***genome_annotations*** table that is saved in the data directory. Note that the airtable connection requires a personal access token. 

```{r download_genome_annotations, warning=FALSE, comments="", message=FALSE, eval=FALSE}
airtable("MAGs", "appWbHBNLE6iAsMRV") %>% #get base ID from Airtable browser URL
  read_airtable(., fields = c("ID","mag_name","number_genes","anno_url"), id_to_col = TRUE) %>% #get 3 columns from MAGs table
  filter(mag_name %in% paste0(genome_metadata$genome,".fa")) %>% #filter by MAG name
  filter(number_genes > 0) %>% #genes need to exist
  select(anno_url) %>% #list MAG annotation urls
  pull() %>%
  read_tsv() %>% #load all tables
  rename(gene=1, genome=2, contig=3) %>% #rename first 3 columns
  write_tsv(file="data/genome_annotations.tsv.xz") #write to overall compressed file
```

```{r load_genome_annotations, warning=FALSE, comments="", message=FALSE}
genome_annotations <- read_tsv("data/genome_annotations.tsv.xz") %>%
    rename(gene=1, genome=2, contig=3)
```

## Create working objects

Transform the original data files into working objects for downstream analyses.

### Filter reads by coverage

```{r filter_coverage, warning=FALSE, comments="", message=FALSE}
min_coverage=0.3
read_counts_filt <- genome_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * read_counts[[cur_column()]])) 
```

### Transform reads into genome counts

```{r calculate_genome_counts_unfiltered, warning=FALSE, comments="", message=FALSE}
readlength=150
genome_counts <- read_counts %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

```{r calculate_genome_counts_filtered, warning=FALSE, comments="", message=FALSE}
readlength=150
genome_counts_filt <- read_counts_filt %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))
```

### Distill annotations into GIFTs 

```{r distill_annotations, warning=FALSE, comments="", message=FALSE, eval=FALSE}
genome_gifts <- distill(genome_annotations,GIFT_db,genomecol=2,annotcol=c(9,10,19))
```

## Prepare color scheme

[AlberdiLab](www.alberdilab.dk) projects use unified color schemes developed for the [Earth Hologenome Initiative](www.earthhologenome.org), to facilitate figure interpretation.

```{r get_ehi_colors, warning=FALSE, comments="", message=FALSE}
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>% 
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)
```

## Wrap working objects

All working objects are wrapped into a single Rdata object to facilitate downstream usage.

```{r wrap_working_objects, warning=FALSE, comments="", message=FALSE}
save(sample_metadata, 
     genome_metadata, 
     read_counts, 
     genome_counts, 
     genome_counts_filt, 
     genome_tree,
     genome_gifts, 
     phylum_colors,
     file = "data/data.Rdata")
```

<!--chapter:end:01_data_preparation.Rmd-->

# Data statistics

```{r load_data_stats}
load("data/data.Rdata")
extra_metadata <- read_csv("data/Bombina_microbiome_metadata.csv") %>% 
  rename(animal=sample) %>% 
  rename(sample=EHI_code)
```
```{r}
filtered_reads <- read_counts %>% 
  column_to_rownames(., "genome") %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rename("depth"=".") %>% 
  rownames_to_column(., "sample") %>% 
  left_join(extra_metadata, by="sample") %>% 
  arrange(animal) %>% 
  group_by(animal) %>%
  slice_max(order_by = depth, n = 1, with_ties = FALSE)
```

## Sequencing reads statistics

```{r reads_stats}
sample_metadata %>%
  summarise(Total=sum(reads_post_fastp * 150 / 1000000000) %>% round(2), 
            mean=mean(reads_post_fastp * 150 / 1000000000) %>% round(2),
            sd=sd(reads_post_fastp * 150 / 1000000000) %>% round(2)) %>%
  unite("Average",mean, sd, sep = " ± ", remove = TRUE) %>%
  tt()
```


## DNA fractions
```{r dna_fractions_stats}
sequence_fractions <- read_counts %>%
    select(all_of(c("genome",filtered_reads$sample))) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "value") %>%
  group_by(sample) %>%
  summarise(mags = sum(value)) %>%
	left_join(sample_metadata, by = join_by(sample == sample)) %>%
	select(sample,mags,metagenomic_bases,host_bases,bases_lost_fastp_percent) %>%
	mutate(mags_bases = mags*146) %>%
	mutate(lowqual_bases = ((metagenomic_bases+host_bases)/(1-bases_lost_fastp_percent))-(metagenomic_bases+host_bases)) %>%
	mutate(unmapped_bases = metagenomic_bases - mags_bases) %>%
	mutate(unmapped_bases = ifelse(unmapped_bases < 0, 0, unmapped_bases)) %>%
	select(sample, lowqual_bases, host_bases, unmapped_bases, mags_bases)

sequence_fractions %>%
  mutate_at(vars(-sample), ~./1000000000) %>%
  rename("Sample"=1, "Low quality"=2, "Mapped to host"=3, "Unmapped"=4, "Mapped to MAGs"=5) %>%
  tt()
```


```{r dna_fractions_plot, message=FALSE, warning=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
sequence_fractions %>%
	pivot_longer(!sample, names_to = "fraction", values_to = "value") %>%
  left_join(., extra_metadata, by="sample") %>% 
	mutate(value = value / 1000000000) %>%
	mutate(fraction = factor(fraction, levels = c("lowqual_bases","host_bases","unmapped_bases","mags_bases"))) %>%
	ggplot(., aes(x = sample, y = value, fill=fraction)) +
  geom_bar(position="stack", stat = "identity") +
  scale_fill_manual(name="Sequence type",
                    breaks=c("lowqual_bases","host_bases","unmapped_bases","mags_bases"),
                    labels=c("Low quality","Mapped to host","Unmapped","Mapped to MAGs"),
                    values=c("#CCCCCC", "#bcdee1", "#d8b8a3","#93655c"))+
  facet_grid(~ Bd_status, scale="free", space="free")+
  labs(x = "Samples", y = "Amount of data (GB)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6),legend.position = "bottom")
```

## Recovered microbial fraction

```{r data_estimations_plot, message=FALSE, warning=FALSE, fig.height=12, fig.width=10, fig.fullwidth=TRUE}
singlem_table <- sequence_fractions %>%
	mutate(mags_proportion = round((mags_bases / (mags_bases + unmapped_bases))*100,2)) %>%
	left_join(sample_metadata, by = join_by(sample == sample))  %>%
	mutate(singlem_proportion = round(singlem_fraction*100,2)) %>%
	select(sample,mags_proportion,singlem_proportion) %>%
	mutate(mags_proportion = ifelse(singlem_proportion == 0, 0, mags_proportion)) %>% #convert zeros to NA
	mutate(singlem_proportion = ifelse(singlem_proportion == 0, NA, singlem_proportion)) %>% #convert zeros to NA
	mutate(singlem_proportion = ifelse(singlem_proportion < mags_proportion, NA, singlem_proportion)) %>% #if singlem is smaller, then NA, to simplify plot
	mutate(singlem_proportion = ifelse(singlem_proportion > 100, 100, singlem_proportion)) #simplify

singlem_table %>%
	pivot_longer(!sample, names_to = "proportion", values_to = "value") %>%
	left_join(sample_metadata, by = join_by(sample == sample))  %>%
	mutate(proportion = factor(proportion, levels = c("mags_proportion","singlem_proportion"))) %>%
	ggplot(., aes(x = value, y = sample, color=proportion)) +
			geom_line(aes(group = sample), color = "#f8a538") +
			geom_point() +
      scale_color_manual(name="Proportion",
                    breaks=c("mags_proportion","singlem_proportion"),
                    labels=c("Recovered","Estimated"),
                    values=c("#52e1e8", "#876b53"))+
      facet_nested(species + sample_type ~ ., scales="free",space="free")+
			theme_classic() +
			labs(y = "Samples", x = "Prokaryotic fraction (%)") +
	    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size=6),
	          legend.position = "right",
	          strip.background.y=element_rect(color = NA, fill= "#f4f4f4"))
```




<!--chapter:end:02_data_statistics.Rmd-->

# MAG catalogue

```{r load_data_mag, message=FALSE, warning=FALSE}
load("data/data.Rdata")
```

## Genome phylogeny

```{r genome_phylogeny, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}
# Generate the phylum color heatmap
phylum_heatmap <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(genome,phylum) %>%
    mutate(phylum = factor(phylum, levels = unique(phylum))) %>%
    column_to_rownames(var = "genome")

# Generate  basal tree
circular_tree <- force.ultrametric(genome_tree, method="extend") %>% # extend to ultrametric for the sake of visualisation
    ggtree(., layout="fan", open.angle=10, size=0.5)

# Add phylum ring
circular_tree <- gheatmap(circular_tree, phylum_heatmap, offset=0.55, width=0.1, colnames=FALSE) +
        scale_fill_manual(values=phylum_colors) +
        geom_tiplab2(size=1, hjust=-0.1) +
        theme(legend.position = "none", plot.margin = margin(0, 0, 0, 0), panel.margin = margin(0, 0, 0, 0))

# Flush color scale to enable a new color scheme in the next ring
circular_tree <- circular_tree + new_scale_fill()

# Add completeness ring
circular_tree <- circular_tree +
        new_scale_fill() +
        scale_fill_gradient(low = "#d1f4ba", high = "#f4baba") +
        geom_fruit(
                data=genome_metadata,
                geom=geom_bar,
                mapping = aes(x=completeness, y=genome, fill=contamination),
                offset = 0.55,
                orientation="y",
              stat="identity")

# Add genome-size ring
circular_tree <-  circular_tree +
        new_scale_fill() +
        scale_fill_manual(values = "#cccccc") +
        geom_fruit(
             data=genome_metadata,
             geom=geom_bar,
             mapping = aes(x=length, y=genome),
                 offset = 0.05,
                 orientation="y",
         stat="identity")

# Add text
circular_tree <-  circular_tree +
        annotate('text', x=2.7, y=0, label='            Phylum', family='arial', size=3.5) +
        annotate('text', x=3.1, y=0, label='                         Genome quality', family='arial', size=3.5) +
        annotate('text', x=3.5, y=0, label='                     Genome size', family='arial', size=3.5)

#Plot circular tree
circular_tree %>% open_tree(30) %>% rotate_tree(90)
```

## Genome quality

```{r genome_quality}
genome_metadata %>% 
    summarise(completeness_mean=mean(completeness) %>% round(2) %>% as.character(), 
              completeness_sd=sd(completeness) %>% round(2) %>% as.character(), 
              contamination_mean=mean(contamination) %>% round(2), 
              contamination_sd=sd(contamination) %>% round(2)) %>%
    unite("Completeness",completeness_mean, completeness_sd, sep = " ± ", remove = TRUE) %>%
    unite("Contamination",contamination_mean, contamination_sd, sep = " ± ", remove = TRUE) %>%
    tt()
```

```{r genome_quality_plot, message=FALSE, warning=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}

#Generate quality biplot
genome_biplot <- genome_metadata %>%
  select(c(genome,domain,phylum,completeness,contamination,length)) %>%
  arrange(match(genome, rev(genome_tree$tip.label))) %>% #sort MAGs according to phylogenetic tree
  ggplot(aes(x=completeness,y=contamination,size=length,color=phylum)) +
              geom_point(alpha=0.7) +
                    ylim(c(10,0)) +
                    scale_color_manual(values=phylum_colors) +
                    labs(y= "Contamination", x = "Completeness") +
                    theme_classic() +
                    theme(legend.position = "none")

#Generate contamination boxplot
genome_contamination <- genome_metadata %>%
            ggplot(aes(y=contamination)) +
                    ylim(c(10,0)) +
                    geom_boxplot(colour = "#999999", fill="#cccccc") +
                    theme_void() +
                    theme(legend.position = "none",
                        axis.title.x = element_blank(),
                        axis.title.y = element_blank(),
                        axis.text.y=element_blank(),
                        axis.ticks.y=element_blank(),
                        axis.text.x=element_blank(),
                        axis.ticks.x=element_blank(),
                        plot.margin = unit(c(0, 0, 0.40, 0),"inches")) #add bottom-margin (top, right, bottom, left)

#Generate completeness boxplot
genome_completeness <- genome_metadata %>%
        ggplot(aes(x=completeness)) +
                xlim(c(50,100)) +
                geom_boxplot(colour = "#999999", fill="#cccccc") +
                theme_void() +
                theme(legend.position = "none",
                    axis.title.x = element_blank(),
                    axis.title.y = element_blank(),
                    axis.text.y=element_blank(),
                    axis.ticks.y=element_blank(),
                    axis.text.x=element_blank(),
                    axis.ticks.x=element_blank(),
                    plot.margin = unit(c(0, 0, 0, 0.50),"inches")) #add left-margin (top, right, bottom, left)

#Render composite figure
grid.arrange(grobs = list(genome_completeness,genome_biplot,genome_contamination),
        layout_matrix = rbind(c(1,1,1,1,1,1,1,1,1,1,1,4),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3),
                              c(2,2,2,2,2,2,2,2,2,2,2,3)))

```


## Functional overview

```{r function_heatmap, message=FALSE, warning=FALSE, fig.height=10, fig.width=10, fig.fullwidth=TRUE}

# Aggregate basal GIFT into elements
function_table <- genome_gifts %>%
    to.elements(., GIFT_db)

# Generate  basal tree
function_tree <- force.ultrametric(genome_tree, method="extend") %>%
                ggtree(., size = 0.3) 

#Add phylum colors next to the tree tips
function_tree <- gheatmap(function_tree, phylum_heatmap, offset=0, width=0.1, colnames=FALSE) +
            scale_fill_manual(values=phylum_colors) +
            labs(fill="Phylum")

#Reset fill scale to use a different colour profile in the heatmap
function_tree <- function_tree + new_scale_fill()

#Add functions heatmap
function_tree <- gheatmap(function_tree, function_table, offset=0.5, width=3.5, colnames=FALSE) +
            vexpand(.08) +
            coord_cartesian(clip = "off") +
            scale_fill_gradient(low = "#f4f4f4", high = "steelblue", na.value="white") +
            labs(fill="GIFT")

#Reset fill scale to use a different colour profile in the heatmap
function_tree <- function_tree + new_scale_fill()

# Add completeness barplots
function_tree <- function_tree +
            geom_fruit(data=genome_metadata,
            geom=geom_bar,
            grid.params=list(axis="x", text.size=2, nbreak = 1),
            axis.params=list(vline=TRUE),
            mapping = aes(x=length, y=genome, fill=completeness),
                 offset = 3.8,
                 orientation="y",
                 stat="identity") +
            scale_fill_gradient(low = "#cf8888", high = "#a2cc87") +
            labs(fill="Genome\ncompleteness")

function_tree
```

## Functional ordination

```{r function_ordination, message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE, eval=FALSE}
# Generate the tSNE ordination
tSNE_function <- Rtsne(X=function_table, dims = 2, check_duplicates = FALSE)

# Plot the ordination
function_ordination <- tSNE_function$Y %>%
                as.data.frame() %>%
                mutate(genome=rownames(function_table)) %>%
                inner_join(genome_metadata, by="genome") %>%
                rename(tSNE1="V1", tSNE2="V2") %>%
                select(genome,phylum,tSNE1,tSNE2, length) %>%
                ggplot(aes(x = tSNE1, y = tSNE2, color = phylum, size=length))+
                            geom_point(shape=16, alpha=0.7) +
                            scale_color_manual(values=phylum_colors) +
                            theme_minimal() +
                labs(color="Phylum", size="Genome size") +
                guides(color = guide_legend(override.aes = list(size = 5))) # enlarge Phylum dots in legend

function_ordination
```

<!--chapter:end:03_mag_catalogue.Rmd-->

# Compositional analysis

```{r load_data_composition}
load("data/data.Rdata")
extra_metadata <- read_csv("data/Bombina_microbiome_metadata.csv") %>% 
  rename(animal=sample) %>% 
  rename(sample=EHI_code)
sample_metadata <- sample_metadata %>% 
  left_join(., extra_metadata, by="sample")

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome)

phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>% 
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)
```

```{r}
sample_metadata %>% 
  group_by(region, Bd_status, sex) %>% 
  summarise(count=n())
```

### Taxonomy barplot per individual

```{r barplot_individuals, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(!is.na(count)) %>%
  ggplot(aes(y=count,x=sample, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + 
    scale_fill_manual(values=phylum_colors) +
  facet_grid(~ Bd_status, scale="free", space="free")+
    guides(fill = guide_legend(ncol = 1)) +
    theme(
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 8, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```

## Taxonomy boxplot
### Phylum relative abundances

```{r taxonomy_phylum_summary, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,Bd_status) %>%
  summarise(relabun=sum(count))
```

#### Origin: Dogs vs Sled dogs vs Wolves
```{r taxonomy_phylum_summary_origin, warning=FALSE, comments="", message=FALSE}
phylum_summary %>%
  group_by(phylum) %>%
  summarise(
    total_mean = mean(relabun * 100, na.rm = T),
    total_sd = sd(relabun * 100, na.rm = T),
    positive_mean = mean(relabun[Bd_status == "Bd_positive"] * 100, na.rm = T),
    positive_sd = sd(relabun[Bd_status == "Bd_positive"] * 100, na.rm = T),
    negative_mean = mean(relabun[Bd_status == "Bd_negative"] * 100, na.rm = T),
    negative_sd = sd(relabun[Bd_status == "Bd_negative"] * 100, na.rm = T)
  ) %>%
  mutate(
    total = str_c(round(total_mean, 3), "±", round(total_sd, 3)),
    Positive = str_c(round(positive_mean, 3), "±", round(positive_sd, 3)),
    Negative = str_c(round(negative_mean, 3), "±", round(negative_sd, 3))
    ) %>%
  arrange(-total_mean) %>%
  dplyr::select(phylum, total, Positive, Negative) %>%
  tt()
```





<!--chapter:end:04_microbiome_composition.Rmd-->

# Differential abundance analysis
```{r load_data_dif}
set.seed(1234) 
load("data/data.Rdata")
extra_metadata <- read_csv("data/Bombina_microbiome_metadata.csv") %>% 
  rename(animal=sample) %>% 
  rename(sample=EHI_code)
sample_metadata <- sample_metadata %>% 
  left_join(., extra_metadata, by="sample")

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome)

phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>% 
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)

count_phy <- genome_counts %>%
  column_to_rownames(var="genome")%>%
  otu_table(., taxa_are_rows=T)

sample_info_tab_phy <- sample_metadata%>%
  column_to_rownames(var="sample")%>%
  sample_data()

TAX <- genome_metadata%>%
  column_to_rownames(var="genome")%>%
  select(1:7)%>%
  as.matrix()%>%
  tax_table()

tree <- phy_tree(genome_tree)

physeq_all = phyloseq(count_phy, TAX, sample_info_tab_phy, tree)
```

## Ancombc2

### MAG level
```{r ancom_rand_Bd, comment="", message=FALSE, warning=FALSE, eval=FALSE}
ancom_rand_output_mag = ancombc2(data = physeq_all, 
                  assay_name = "counts", 
                  tax_level = NULL, 
                  fix_formula = "Bd_status", 
                  p_adj_method = "BH", 
                  pseudo_sens = TRUE,
                  prv_cut =0, 
                  lib_cut = 0, 
                  s0_perc = 0.05,
                  group = "Bd_status", 
                  struc_zero = TRUE, 
                  neg_lb = FALSE,
                  alpha = 0.05, 
                  n_cl = 2, 
                  verbose = TRUE,
                  global = FALSE, 
                  pairwise = FALSE, 
                  dunnet = FALSE, 
                  trend = FALSE,
                  iter_control = list(tol = 1e-5, max_iter = 20, verbose = FALSE),
                  em_control = list(tol = 1e-5, max_iter = 100),
                  mdfdr_control = list(fwer_ctrl_method = "holm", B = 100), 
                  trend_control = NULL)
```

**Total number of structural zeros**
```{r}
# Extract structural zeros
structural_zeros <- ancom_rand_output_mag$zero_ind

# Convert to a data frame for better readability
structural_zeros_df <- as.data.frame(structural_zeros) %>% 
  column_to_rownames(., "taxon")

rownames(structural_zeros_df)[apply(structural_zeros_df, 1, any)]
```

**Enriched taxa**
```{r ancom_rand, comment="", message=FALSE, warning=FALSE}
ancom_rand_output_mag$res %>%
  dplyr::select(taxon, lfc_Bd_statusBd_positive, q_Bd_statusBd_positive) %>%
  filter(q_Bd_statusBd_positive < 0.05)
```

<!--chapter:end:05_differential_abundance.Rmd-->

# Alpha diversity

```{r load_data_diversity}
set.seed(1234) 

load("data/data.Rdata")

extra_metadata <- read_csv("data/Bombina_microbiome_metadata.csv") %>% 
  rename(animal=sample) %>% 
  rename(sample=EHI_code)
sample_metadata <- sample_metadata %>% 
  left_join(., extra_metadata, by="sample")

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome)

phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>% 
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)
```
```{r alpha_div, comment="", message=FALSE, warning=FALSE}
richness <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 0) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(richness = 1) %>%
  rownames_to_column(var = "sample")

neutral <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(neutral = 1) %>%
  rownames_to_column(var = "sample")

phylogenetic <- genome_counts_filt %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, tree = genome_tree) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(phylogenetic = 1) %>%
  rownames_to_column(var = "sample")

# Aggregate basal GIFT into elements
dist <- genome_gifts %>%
  to.elements(., GIFT_db) %>%
  traits2dist(., method = "gower")

functional <- genome_counts_filt %>%
  filter(genome %in% labels(dist)[[1]]) %>%
  column_to_rownames(var = "genome") %>%
  dplyr::select(where(~ !all(. == 0))) %>%
  hilldiv(., q = 1, dist = dist) %>%
  t() %>%
  as.data.frame() %>%
  dplyr::rename(functional = 1) %>%
  rownames_to_column(var = "sample") %>%
  mutate(functional = if_else(is.nan(functional), 1, functional))

# Merge all metrics
alpha_div <- richness %>%
  full_join(neutral, by = join_by(sample == sample)) %>%
  full_join(phylogenetic, by = join_by(sample == sample)) %>%
  full_join(functional, by = join_by(sample == sample))
```
```{r alpha_plots_loc, fig.height=5, fig.width=12, fig.fullwidth=TRUE}
alpha_div %>%
  pivot_longer(-sample, names_to = "metric", values_to = "value") %>%
  mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic", "functional"))) %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  ggplot(aes(y = value, x = Bd_status, group=Bd_status, color=Bd_status, fill=Bd_status)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  facet_wrap(. ~factor(metric, labels=c("richness" = "Richness", "neutral" = "Neutral", "phylogenetic" = "Phylogenetic", "functional"="Functional")), scales = "free", ncol = 5) +
  stat_compare_means(show.legend = FALSE) +
  coord_cartesian(xlim = c(1, NA)) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size =12, lineheight = 0.6),
    panel.grid.minor.x = element_line(size = .1, color = "grey"),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.text.x = element_blank()
  )
```

# Beta diversity

```{r beta_div, comment="", message=FALSE, warning=FALSE, eval=FALSE}
beta_q0n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 0)

beta_q1n <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1)

beta_q1p <- genome_counts_filt %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, tree = genome_tree)

beta_q1f <- genome_counts_filt %>%
  filter(genome %in% labels(dist)[[1]]) %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0)) %>%
  hillpair(., q = 1, dist = dist)
```

## Richness
```{r beta_div_nmds_richness, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
beta_q0n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  mutate(Bd_status=factor(Bd_status,levels=c("Bd_positive", "Bd_negative")))%>%
  group_by(Bd_status) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = Bd_status)) +
    geom_point(size = 3, alpha = 0.9) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Bd status')
```

```{r permanova__q0, comment="", message=FALSE, warning=FALSE}
beta_q0n_mat <- as.matrix(beta_q0n$S)
beta_q0n_filtered <- beta_q0n_mat[rownames(beta_q0n_mat) %in% sample_metadata$sample,
                        colnames(beta_q0n_mat) %in% sample_metadata$sample]
beta_q0n_dis <- as.dist(beta_q0n_filtered)

sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q0n_dis))) %>% 
  column_to_rownames(., "sample")

all(sample_metadata_ordered$sample == labels(beta_q0n_dis))

betadisper(beta_q0n_dis, sample_metadata_ordered$Bd_status) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q0n_dis ~ Bd_status+region+sex, 
        data = sample_metadata_ordered, 
        permutations = 999, by="margin") %>%
        broom::tidy() %>%
        tt()
```

## Neutral
```{r beta_div_nmds_neutral, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
beta_q1n$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  mutate(Bd_status=factor(Bd_status,levels=c("Bd_positive", "Bd_negative")))%>%
  group_by(Bd_status) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = Bd_status)) +
    geom_point(size = 3, alpha = 0.9) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Bd status')
```

```{r permanova__q1, comment="", message=FALSE, warning=FALSE}
beta_q1n_mat <- as.matrix(beta_q1n$S)
beta_q1n_filtered <- beta_q1n_mat[rownames(beta_q1n_mat) %in% sample_metadata$sample,
                        colnames(beta_q1n_mat) %in% sample_metadata$sample]
beta_q1n_dis <- as.dist(beta_q1n_filtered)

sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1n_dis))) %>% 
  column_to_rownames(., "sample")

all(sample_metadata_ordered$sample == labels(beta_q1n_dis))

betadisper(beta_q1n_dis, sample_metadata_ordered$Bd_status) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1n_dis ~ Bd_status+region+sex, 
        data = sample_metadata_ordered, 
        permutations = 999, by="margin") %>%
        broom::tidy() %>%
        tt()
```

## Phylogenetic
```{r beta_div_nmds_phylo, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
beta_q1p$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  mutate(Bd_status=factor(Bd_status,levels=c("Bd_positive", "Bd_negative")))%>%
  group_by(Bd_status) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = Bd_status)) +
    geom_point(size = 3, alpha = 0.9) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Bd status')
```
```{r permanova__q1p, comment="", message=FALSE, warning=FALSE}
beta_q1p_mat <- as.matrix(beta_q1p$S)
beta_q1p_filtered <- beta_q1p_mat[rownames(beta_q1p_mat) %in% sample_metadata$sample,
                        colnames(beta_q1p_mat) %in% sample_metadata$sample]
beta_q1p_dis <- as.dist(beta_q1p_filtered)

sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1p_dis))) %>% 
  column_to_rownames(., "sample")

all(sample_metadata_ordered$sample == labels(beta_q1p_dis))

betadisper(beta_q1p_dis, sample_metadata_ordered$Bd_status) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1p_dis ~ Bd_status+region+sex, 
        data = sample_metadata_ordered, 
        permutations = 999, by="margin") %>%
        broom::tidy() %>%
        tt()
```

## Functional
```{r beta_div_nmds_func, fig.height=7, fig.width=12, fig.fullwidth=TRUE}
beta_q1f$S %>%
  vegan::metaMDS(., trymax = 500, k = 2, trace=0) %>%
  vegan::scores() %>%
  as_tibble(., rownames = "sample") %>%
  dplyr::left_join(sample_metadata, by = join_by(sample == sample)) %>%
  mutate(Bd_status=factor(Bd_status,levels=c("Bd_positive", "Bd_negative")))%>%
  group_by(Bd_status) %>%
  mutate(x_cen = mean(NMDS1, na.rm = TRUE)) %>%
  mutate(y_cen = mean(NMDS2, na.rm = TRUE)) %>%
  ungroup() %>% 
  ggplot(aes(x = NMDS1, y = NMDS2, color = Bd_status)) +
    geom_point(size = 3, alpha = 0.9) +
    geom_segment(aes(x = x_cen, y = y_cen, xend = NMDS1, yend = NMDS2), alpha = 0.5, show.legend = FALSE) +
    theme_classic() +
    theme(
      axis.text.x = element_text(size = 12),
      axis.text.y = element_text(size = 12),
      axis.title = element_text(size = 12, face = "bold"),
      axis.text = element_text(face = "bold", size = 12),
      panel.background = element_blank(),
      axis.line = element_line(size = 0.5, linetype = "solid", colour = "black"),
      legend.text = element_text(size = 12),
      legend.title = element_text(size = 14),
      legend.position = "right", legend.box = "vertical"
    ) +
  labs(color='Bd status')
```

```{r permanova__q1f, comment="", message=FALSE, warning=FALSE}
beta_q1f_mat <- as.matrix(beta_q1f$S)
beta_q1f_filtered <- beta_q1f_mat[rownames(beta_q1f_mat) %in% sample_metadata$sample,
                        colnames(beta_q1f_mat) %in% sample_metadata$sample]
beta_q1f_dis <- as.dist(beta_q1f_filtered)

sample_metadata_ordered <- sample_metadata %>% 
  arrange(match(sample,labels(beta_q1f_dis))) %>% 
  column_to_rownames(., "sample")

all(sample_metadata_ordered$sample == labels(beta_q1f_dis))

betadisper(beta_q1f_dis, sample_metadata_ordered$Bd_status) %>% permutest(., pairwise = TRUE) 

adonis2(beta_q1f_dis ~ Bd_status+region+sex, 
        data = sample_metadata_ordered, 
        permutations = 999, by="margin") %>%
        broom::tidy() %>%
        tt()
```

<!--chapter:end:06_diversity.Rmd-->

# Functional differences
```{r load_data_funct}
set.seed(1234) 

load("data/data.Rdata")

extra_metadata <- read_csv("data/Bombina_microbiome_metadata.csv") %>% 
  rename(animal=sample) %>% 
  rename(sample=EHI_code)
sample_metadata <- sample_metadata %>% 
  left_join(., extra_metadata, by="sample")

genome_counts_filt <- genome_counts_filt %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))

genome_metadata <- genome_metadata %>%
  semi_join(., genome_counts_filt, by = "genome") %>%
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome)

phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
    right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    arrange(match(genome, genome_tree$tip.label)) %>%
    select(phylum, colors) %>% 
    unique() %>%
    arrange(phylum) %>%
    pull(colors, name=phylum)
```
```{r gift_analysis_func, comment="", message=FALSE, warning=FALSE}
genome_counts_filt <- genome_counts_filt[genome_counts_filt$genome %in% rownames(genome_gifts),]
genome_counts_filt <- genome_counts_filt %>%
    select(genome, all_of(sample_metadata$sample)) %>%
  column_to_rownames(., "genome") %>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))%>%
  rownames_to_column(., "genome")

genome_gifts <- genome_gifts[rownames(genome_gifts) %in% genome_counts_filt$genome,]
genome_gifts <- genome_gifts[, colSums(genome_gifts != 0) > 0]

#Aggregate bundle-level GIFTs into the compound level
GIFTs_elements <- to.elements(genome_gifts,GIFT_db)
GIFTs_elements_filtered <- GIFTs_elements[rownames(GIFTs_elements) %in% genome_counts_filt$genome,]
GIFTs_elements_filtered <- as.data.frame(GIFTs_elements_filtered) %>% 
  select_if(~ !is.numeric(.) || sum(.) != 0)

#Aggregate element-level GIFTs into the function level
GIFTs_functions <- to.functions(GIFTs_elements_filtered,GIFT_db)

#Aggregate function-level GIFTs into overall Biosynthesis, Degradation and Structural GIFTs
GIFTs_domains <- to.domains(GIFTs_functions,GIFT_db)

#Get community-weighed average GIFTs per sample
genome_counts_row <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% 
  column_to_rownames(., "genome") 
GIFTs_elements_community <- to.community(GIFTs_elements_filtered,genome_counts_row,GIFT_db)
GIFTs_functions_community <- to.community(GIFTs_functions,genome_counts_row,GIFT_db)
GIFTs_domains_community <- to.community(GIFTs_domains,genome_counts_row,GIFT_db)
```

## Bd_status

```{r gitfs_functional_bd, comment="", message=FALSE, warning=FALSE}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(Bd_status) %>%
  summarise(MCI = mean(value), sd = sd(value)) %>%
  tt()
```


```{r commun_wilcox_elem_bd, comment="", message=FALSE, warning=FALSE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata %>% select(sample,Bd_status), by="sample")

uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

element_gift %>%
    pivot_longer(-c(sample,Bd_status), names_to = "trait", values_to = "value") %>%
    group_by(trait) %>%
    summarise(p_value = wilcox.test(value ~ Bd_status)$p.value) %>%
    mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
    filter(p_adjust < 0.05)
```

## region

```{r gitfs_functional_region, comment="", message=FALSE, warning=FALSE}
GIFTs_functions_community %>%
  rowMeans() %>%
  as_tibble(., rownames = "sample") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  group_by(region) %>%
  summarise(MCI = mean(value), sd = sd(value)) %>%
  tt()
```

```{r commun_wilcox_elem_region, comment="", message=FALSE, warning=FALSE}
element_gift <- GIFTs_elements_community %>% 
  as.data.frame() %>% 
  rownames_to_column(., "sample") %>% 
  left_join(sample_metadata %>% select(sample,region), by="sample")

uniqueGIFT_db<- unique(GIFT_db[c(2,4,5,6)]) %>% unite("Function",Function:Element, sep= "_", remove=FALSE)

element_gift %>%
    pivot_longer(-c(sample,region), names_to = "trait", values_to = "value") %>%
    group_by(trait) %>%
    summarise(p_value = wilcox.test(value ~ region)$p.value) %>%
    mutate(p_adjust=p.adjust(p_value, method="BH")) %>%
    filter(p_adjust < 0.05)
```

<!--chapter:end:07_functional_differences.Rmd-->

