---
title: "AlberdiLab | van-Riemsdijk et al. in prep"
subtitle: Comparison of chytrid infected (Bd positive) and uninfected (Bd negative) Bombina individuals from natural populations.

date: "`r Sys.Date()`"

authors:
- Isolde van Riemsdijk^[University of Copenhagen, isolde.vanriemsdijk@sund.ku.dk]
- Ostaizka Aizpurua^[University of Copenhagen, ostaizka.aizpurua@sund.ku.dk]
- Antton Alberdi^[University of Copenhagen, antton.alberdi@sund.ku.dk]

site: bookdown::bookdown_site

documentclass: book

bibliography:
- book.bib
- packages.bib

url: https://alberdilab.github.io/bombina_hologenomics

description: |
  Data analysis code for the study on the recovery of metagenome‑assembled genomes and derived microbial communities from Bombina toad skin samples, with Bd infected and healthy individuals.

link-citations: true

github-repo: alberdilab/bombina_hologenomics
---

```{r knitr_opts, echo=FALSE}
knitr::opts_chunk$set(
    class.source = "script-source",
    class.output = "script-output",
    comment = NA)
```

# Introduction

This webbook contains all the code used for data analysis in study of the population-level metagenomic data of Podarcis muralis lizards across elevational gradients in various mountain ranges of the Pyrenees.

## Prepare the R environment

### Environment

To reproduce all the analyses locally, clone this repository in your computer using:

```
RStudio > New Project > Version Control > Git
```

And indicating the following git repository:

> https://github.com/alberdilab/bombina_hologenomics.git

Once the R project has been created, follow the instructions and code chunks shown in this webbook.

### Libraries

The following R packages are required for the data analysis.

```{r load_libraries, warning=TRUE, comments="", message=TRUE}
# Base
library(R.utils)
library(knitr)
library(tidyverse)
library(devtools)
library(tinytable)
library(rairtable)
library(stringr)

# For tree handling
library(ape)
library(phyloseq) 
library(phytools)

# For plotting
library(ggplot2)
library(ggrepel)
library(ggpubr)
library(ggnewscale)
library(gridExtra)
library(ggtree)
library(ggtreeExtra)
library(ggh4x)

# For statistics
library(nlme)
library(spaa)
library(vegan)
library(Rtsne)
library(geiger)
library(hilldiv2)
library(distillR)
library(broom.mixed)
library(corrplot)
library(ANCOMBC) ## package not available for version 4.5
library(microbiome)
library(lmerTest)
```

### Installation trouble shooting
If some of the above packages are not installed use:

BiocManager:
phyloseq
gridExtra
ggtreeExtra

Devtools & github:
ggtree (GuangchuangYu/ggtree)
hilldiv2 (anttonalberdi/hilldiv2)
distillR (anttonalberdi/distillR)
ANCOMBC (FrederickHuangLin/ANCOMBC)

```{r install_libs, eval=F}
# For BiocManager use:
if (!require("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install("phyloseq")

# For Github packages use:
pak::pak("GuangchuangYu/ggtree")
```
